coauthor_comments

### 2021-04-20 doc/manuscript/ForestGEO_treerings.Rmd
worked on coauthor feedback

\newpage
**Table 1 | (DRAFT hypothesis table-- needs to be filled in and condensed/ cleaned up)**. .... 
```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(flextable)
set_flextable_defaults(font.family = "Arial", font.size = 9)
hypotheses_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
hypotheses_table <- subset(hypotheses_table, select = -c(table_row, Q, H, category, notes) )#remove columns that shouldn't be included
ft <- flextable(hypotheses_table)
ft <- italic(ft, i=c(1,15,21), j=1)
ft <- bold(ft, i=c(1,15,21, 2,6,12,16,22))
ft <- autofit(ft)
ft <- footnote( ft, i = 1, j = 2,
                value = as_paragraph(c("SSC= species-site combinations; SSCV=species-site-variable combination")),
                ref_symbols = c("*"),
                part = "header",inline=T)
ft

@@

Growth rate -- whether measured as RW, BAI, or $\Delta$AGB -- varied nonlinearly with DBH for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig. `r size`). 
Variation in these patterns -- particularly for RW, for which variation was most pronounced -- was driven by two primary, interrelated factors: species ecology and stand history.
Species that would have established in fairly open conditions -- i.e., shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC) -- exhibited rapid initial growth, measured as RW, followed by a roughly exponential decline.
Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments [@fritts_tree_1976; @biondi_theorydriven_2008].
However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, RW.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (i.e., a "cross-sectional" analysis), RW increases continuously with DBH [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].

@@

*Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of DBH by yeaar (Fig. `r year_fig`).*
*These results were similar across growth metrics (Figs. `r all_GLS`), and consistent with an independent analysis of growth-DBH trends by decade (Figs. `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.* (**EG: THESE SENTENCES (and the following) SOUND MORE LIKE RESULTS/ METHODS**)
Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach **conceptually?** parallel (**AT comments that it doesn't seem parallel**) to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in RW and BAI with DBH (Fig. `r size`).
For instance, based on our finding that BAI tends to rise and fall with DBH, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
However, our results remain subject to some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig. `r year_fig`, @brienen_detecting_2012; @groenendijk_no_2015; @brienen_tree_2017; @nehrbass-ahles_influence_2014].
While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
The observed trends should therefore be interpreted as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
For instance, ... *give an example/more clear explanation here? (see Camille's feedback)*
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than the primary drivers, indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).

@@

In a few instances, directional changes in growth are likely attributable to abiotic drivers.
In the Czech Republic (ZOF), acid deposition dramatically reduced growth in the late 20th century, with peak influence between the 1970s and the early 1990s [@elling_dendroecological_2009; @samonil_longterm_2008], as captured in our records (Fig. `r decadal_plot_ZOF`), and this influenced linear growth trends of various species differently within the GLS model.
In cases such as this, a non-linear response function to year, or incorporation of data on pollution, would be needed to accurately capture changing growth trends, but that is beyond the scope of the current analysis.
In New Mexico (LT), where growth rates of *Pinus ponderosa* declined, it is possible that the primary climate drivers identified here do not capture all aspects of a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013].
At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020]; we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.

>>>>

\newpage
**Table 1 | (DRAFT hypothesis table-- needs to be filled in and condensed/ cleaned up)**. .... 
```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(flextable)
set_flextable_defaults(font.family = "Arial", font.size = 9)
hypotheses_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
hypotheses_table <- subset(hypotheses_table, select = -c(table_row, Q, H, category, notes) )#remove columns that shouldn't be included
ft <- flextable(hypotheses_table)
ft <- italic(ft, i=c(1,15,21), j=1)
ft <- bold(ft, i=c(1,15,21, 2,6,12,16,22))
ft <- line_spacing(ft, space = 1, part = "all")
ft <- autofit(ft)
ft <- footnote( ft, i = 1, j = 2,
                value = as_paragraph(c("SSC= species-site combinations; SSCV=species-site-variable combination")),
                ref_symbols = c("*"),
                part = "header",inline=T)
ft

@@

Growth rate -- whether measured as RW, BAI, or $\Delta$AGB -- varied nonlinearly with DBH for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig. `r size`). 
Variation in these patterns -- particularly for RW, for which variation was most pronounced -- was driven by two primary, interrelated factors: species ecology and stand history.
Species that would have established in fairly open conditions -- i.e., shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC) -- exhibited rapid initial growth, measured as RW, followed by a roughly exponential decline.
Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments, where declining RW is in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008].
However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, RW.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (i.e., a "cross-sectional" analysis), RW increases continuously with DBH [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].

@@

*Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of DBH by yeaar (Fig. `r year_fig`).*
*These results were similar across growth metrics (Figs. `r all_GLS`), and consistent with an independent analysis of growth-DBH trends by decade (Figs. `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.* (**EG: THESE SENTENCES (and the following) SOUND MORE LIKE RESULTS/ METHODS**)
Indeed, in a comparative analysis of several methods commonly used to detect growth trends, a conceptually parallel approach to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in RW and BAI with DBH (Fig. `r size`).
For instance, based on our finding that BAI tends to rise and fall with DBH, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
However, our results remain subject to some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig. `r year_fig`, @brienen_detecting_2012; @groenendijk_no_2015; @brienen_tree_2017; @nehrbass-ahles_influence_2014].
While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
The observed trends should therefore be interpreted as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
For instance, the oldest trees, which are the only records available for the earliest decades, were the competitive winners, most likely with above-average growth rates [@aubry-kientz_joint_2015], thereby upwardly biasing growth rates in early decades [@groenendijk_no_2015].
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than the primary drivers, indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).

@@

In a few instances, directional changes in growth are likely attributable to abiotic drivers.
At Zofin, size-corrected growth rates were lowest in the 1970s and 1980s (Fig. `r decadal_plot_ZOF`), consistent with other studies from central Europe showing dramatic growth reductions due to acid deposition during this period [@elling_dendroecological_2009; @samonil_longterm_2008]. 
Non-linear trends such as this would be more accurately described by a non-linear response function to year, or incorporation of data on pollution, but that is beyond the scope of the current analysis.
In New Mexico (LT), where growth rates of *Pinus ponderosa* declined, it is possible that the primary climate drivers identified here do not capture all aspects of a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013].
At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020]; we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.

### 2021-04-20 doc/manuscript/ForestGEO_treerings_SI.Rmd
worked on coauthor feedback

Cores were collected outside the ForestGEO plot, but within a contiguous patch of forest.
In 2012 - 2014, cores were collected from live or dead dominant canopy trees with morphological characteristics typical of older trees [@maxwell_declining_2016].
In 2020, additional cores were collected following an ecological study design, wherein all live trees $\ge$ **##** cm within a **(SIZE)** plot were sampled.

@@

The $\v{Z}$of$\'{i}$n forest (ZOF) is a well-preserved natural forest remnant dominated by *Fagus sylvatica* L., *Picea abies* L. and *Abies alba* Mill.
The forest is not fully without past human direct interventions, and a limited historical land use effect is also possible [see @kozakova_contrasting_2011].
Acid deposition dramatically reduced growth in the late 20th century, with peak influence between the 1970s and the early 1990s [@elling_dendroecological_2009; @samonil_long-term_2008].

>>>>

Cores were collected outside the ForestGEO plot, but within a contiguous patch of forest.
In 2012 - 2014, cores were collected from live or dead dominant canopy trees with morphological characteristics typical of older trees [@maxwell_declining_2016].
In 2020, additional cores were collected following an ecological study design, wherein nest subplots were used within a large plot to increase sample depth across age classes of the forest. The first subplot was 5.0 m in radius and all trees larger than 5 cm diameter at breast height (DBH) were cored. In the next subplot, 5-20 m from plot center, all trees >15 cm DBH were cored. In the final subplot, 20-30 m from plot center, all trees >40 cm DBH were cored.

@@

The $\v{Z}$of$\'{i}$n forest (ZOF) is a well-preserved natural forest remnant dominated by *Fagus sylvatica* L., *Picea abies* L. and *Abies alba* Mill.
The forest is not fully without past human direct interventions, and a limited historical land use effect is also possible [see @kozakova_contrasting_2011].

### 2021-04-20 doc/manuscript/coauthor_feedback/2021_02/coauthor feedback.Rmd
Update coauthor feedback.Rmd
worked on coauthor feedback

## BlueCats

@@

@@

@@

@@

@@

## Alan T

@@

## Erika
I think there is very valuable information in the SI (Appendix 2 to 5). For example, there are some details in S4 that would be good to bring to the main text.

@@

## Justin M

@@

@@

## Camille

@@

**Sounds great, and I want to use this alter, but for this paper I can't get it working quickly enough to be worth the time.**

@@

## Alan (re RCS)
I am familiar with RCS, but I don't have much experience using it. One of the main similarities to your method is that it requires sampling across the full age distribution of a species. However, it develops a regional curve of ring width by age, and then this curve is typically used in place of the more common curves (exponential curve or spline) in detrending ring width across the population. Deviations from the regional curve are assumed to represent influences of climate in a particular year or that growing conditions were different than the average for this species (i.e., narrower than average growth up to age 30 could indicate that a tree established under a denser canopy than typical for this species). Your use of DBH and year has some similarities to the use of age in RCS, but DBH is not necessarily strongly correlated with age, and growth trends with calendar year are not necessarily the same as growth trends with age.

@@

>>>>

## BlueCats (DONE)

@@

**Thank you!**

@@

**This would be very interesting, and I'd like to do it, but too much to add to this paper.**

@@

**removed from SI, reworded in main manuscript.**

@@

**[Issue #123](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/123).**

@@

## Alan T (DONE)

@@

**Added mention of this.**

## Erika (DONE)
I think there is very valuable information in the SI (Appendix 2 to 5). For example, there are some details in S4 that would be good to bring to the main text.

**Yes... this feeds into  considerations of how we balance ecology and methods focus. I agree thaat it's valuable info, but unfortunately think it will have to remain a bit buried.**

@@

## Justin M (DONE)

@@

**I'm not completely sure what you mean by connecting to the timing of climate extremes, but at this point would rather just get this done. I am working on bringing out more ecological focus, though.**

@@

## Camille (DONE)

@@

**Much better! Thanks.**

@@

**added**

## Alan (re RCS; DONE)
I am familiar with RCS, but I don't have much experience using it. One of the main similarities to your method is that it requires sampling across the full age distribution of a species. However, it develops a regional curve of ring width by age, and then this curve is typically used in place of the more common curves (exponential curve or spline) in detrending ring width across the population. Deviations from the regional curve are assumed to represent influences of climate in a particular year or that growing conditions were different than the average for this species (i.e., narrower than average growth up to age 30 could indicate that a tree established under a denser canopy than typical for this species). Your use of DBH and year has some similarities to the use of age in RCS, but DBH is not necessarily strongly correlated with age, and growth trends with calendar year are not necessarily the same as growth trends with age.

@@

**reworded a bit**
